---
title: "Working with tabular data"
teaching: 50
exercises: 30
questions:
  - "How to import tabular data into R?"
  - "What kind of object stores tabular data?"
  - "How to investigate the contents of such object?"
  - "How to deal with  missing data?"
objectives: 
  - "Import tabular data into R using the `read_csv()` function."
  - "Understand the relationship between vectors and data.frames/tibbles and distinguish between them."
  - "Discuss the importance of and apply basic quality control checks of data."
  - "Apply the following functions to examine data: `summary()`, `str()`, `unique()`, `length()`"
  - "Analyse vectors with missing data."
keypoints:

source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("02-")
```


## Data frames

`data.frame` is a type of object that contains _tabular_ data. Although these 
can be created "by hand", typically they are the result of reading data from a file. 

There are several functions to read data into R, we're going to use one from the 
`readr` package, which is part of the `tidyverse`:

```{r}
# load the library
library(tidyverse)

# read the data
gapminder2010 <- read_csv("data/gapminder2010_socioeconomic.csv")
```

- import library
- read data
- callout: don't use read.csv
- checklist: specify na values; ensure column types are correct; range of values; missing values
- summary
- str
- subsetting
- 



{% include links.md %}
