---
title: "Manipulating observations (rows) with `dplyr`"
teaching: 50
exercises: 30
questions:

objectives: 
- "introduction to logical comparisons and logical data type"
- "remember and distinguish between different types of logical operators"
- "choose the right conditional operations to obtain specific observations from data using the `filter()` function"
- "apply conditional operations to clean data mistakes and/or highlight elements of interest in graphs using the `ifelse()` function"
- "how to save processed data into a file"
keypoints:

source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("05-")
```

- `filter()`
- how to obtain observations from our data based on certain criteria: `filter()`
  - intro to to logical operators
- Logical operators for missing data
- use of `ifelse` to correct some harder typos "manually"
  - `ifelse(main_religion == "cristian", "christian", main_religion)`
- highlighting observations in graphs with `ifelse()`?
- saving cleaned version of our data: `write_csv()`



If your data include missing values, you may want to become familiar with the
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for
examples.



```{r, purl=FALSE}
## Extract those elements which are not missing values.
rooms[!is.na(rooms)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(rooms)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
rooms[complete.cases(rooms)]
```
